\Large\textbf{Advection versus diffusion in brain ventricular transport}

### 2025-04-02 paper/main.tex
Updates in intro

% I2: What do we know about the problem and the topic?
The flow of CSF in the ventricular system is driven by multiple
physiological processes, including via the beating motion of motile
cilia, CSF secretion by the choroid plexus, and pulsatile pressure gradients generated by the
cardiac cycle, the respiration, neural activity, and bodily
movements~\cite{Brodal2016, del2010ependymal, Ringers2020Role, Vinje2019RespiratoryMeasurements,
Kurtcuoglu2005ComputationalSystem, Olstad2019CiliaryDevelopment,
macaulay2022cerebrospinal, mestre2018flow}.
Cilia are hair-like structures found in a vast diversity of
biological organisms and in several human organs,
of which the brain is one of them~\cite{DGama2025MotileBrain,
Fulford1986Muco-ciliaryLung,
Pacherres2022CiliaryProduction, Jahn1972LocomotionProtozoa,
Brennen1977FluidFlagella, Ringers2020Role, spassky2013motile,
Pellicciotta2020Cilia, mitchell2007evolution,
Tsukita2012CoordinatedFeet, BLAKE1974MechanicsMotion,
Reiten2017Motile-Cilia-MediatedComputations, Thouvenin2020OriginCanal}. 
The ventricular surface of the brain, known as the ependyma, is lined by
motile cilia that beat in a coordinated pattern at frequencies of
10--40 Hz to mediate fluid flow~\cite{mitchell2007evolution,
Ringers2020Role, spassky2013motile, Ringers2023NovelEpithelia, roth2025structure}. 
Given their microscopic size, ranging from
5 to 15 microns, cilia contribute primarily to flow near the
ependyma in the human brain~\cite{Siyahhan2014FlowVentricles, Olstad2019CiliaryDevelopment,
Ringers2020Role, Faubel2016Cilia-basedVentricles}. In contrast, cardiac,
respiratory and delta waves at lower frequencies induce pulsatile flow
at the larger scales in the brain ventricular system, subarachnoid space, and
subarachnoid perivascular
spaces~\cite{Vinje2019RespiratoryMeasurements, eide2024functional,
Causemann2025, fultz2019coupled, mestre2018flow, daversin2020mechanisms}.

% I3: What do we not know about the problem and what do we want to
% know? What are the precise research questions we want to address?
% Refer to specific studies and what specifically we do not know.
To date, it remains poorly understood how motile cilia contribute to 
the overall fluid flow and solute transport within the brain.
Experimental endeavors are challenging to pursue due to
technical limitations: the microscopic size and relatively 
high beating frequencies of the cilia~\cite{ocallaghan2012analysis, mahuzier2018ependymal,
Jeong2022MeasurementTelencephalon, Olstad2019CiliaryDevelopment, spassky2013motile, DGama2021Diversity},
as compared to other fluid dynamic processes in the brain that are slower and span larger
spatial scales~\cite{Vinje2019RespiratoryMeasurements, eide2024functional, fultz2019coupled, mestre2018flow},
make cilia difficult to monitor in vivo in humans.
Until now, most studies have analyzed one physical mechanism at a time, especially in mammals.
For instance, studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles~\cite{Faubel2016Cilia-basedVentricles, Sawamoto2006NewBrain, ibanez2003beat};
however, they failed to identify the role of other contributions to CSF
flow and the accompanying transport, such as pressure
gradients associated with respiration and cardiac pulsations. Moreover, while MRI technology
can measure CSF motion in humans~\cite{feinberg1987human, battal2011cerebrospinal,
yamada2015current}, it cannot detect the microscopic
movements of cilia to understand their impact in the large human
brain~\cite{Siyahhan2014FlowVentricles}. 
These technical limitations of studying cilia-mediated flow and transport make 
mathematical modeling and computational investigations an attractive alternative to experimental
studies. Numerous previous simulation studies have focused on modeling the interaction
between single or several cilia and surrounding
fluids~\cite{Guo2020SimulatingGeometries, Ruvalcaba2021NumericalTree, Smith2009MathematicalFluids,
Cui2019NumericalMethod, Cui2022AFlow}.
Others considered cilia-mediated flow in the brain~\cite{Siyahhan2014FlowVentricles,
Yoshida2022EffectVentricles, Salman2022ComputationalEmbryo, Thouvenin2020OriginCanal}, but also 
other applications such as ciliary transport of mucous and particles in
airways~\cite{Fulford1986Muco-ciliaryLung, Ramirez-SanJuan2020Multi-scaleArrays, roth2025structure},
propulsion and locomotion of ciliated organisms~\cite{BLAKE1974MechanicsMotion,
Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella}, and
oxygen regulation in coral reefs~\cite{Pacherres2022CiliaryProduction}.
These simulation studies have provided valuable insights on how cilia mediate complex
biological flows, and how cilia may contribute to tissue physiology.
However, we still lack a comprehensive understanding of how
cilia contribute to CSF motion and solute transport within the brain.
This is of particular importance since disorders such as hydrocephalus are observed
in animal models and humans with deficient motile cilia~\cite{Afzelius2004CiliaRelatedDiseases,
Olstad2019CiliaryDevelopment, Mitchison2017Motile, Ibanez-Tallon2002LossHydrocephalus},
indicating that the cilia play a crucial role for healthy brain function.

% I4: How do we address the problem and what do we find out?
In this study, we combined computational and experimental approaches to investigate
the role of motile cilia in intraventricular CSF flow and transport.
We model the flow and transport by partial differential equations and solve
them numerically with finite element methods to compute flow patterns and
solute concentration dynamics. A geometry representing the zebrafish larval brain ventricles
was used, for which we have detailed knowledge of cilia properties and CSF motion velocities
and frequencies~\cite{Olstad2019CiliaryDevelopment}. 
Notably, this early developmental stage allows us to simplify our model
by omitting CSF secretion, which starts at later stages~\cite{Jeong2024TheZebrafish},
while still being well-conserved with mammals~\cite{DGama2025MotileBrain,
Olstad2019CiliaryDevelopment, Ringers2020Role, DGama2021Diversity, jurisch2020radial}. 
The computational model is validated with experimental data by monitoring in vivo
the dynamics of a photoconvertible fluorescent protein Dendra2
expressed within the brain ventricles.
Our results show that diffusion plays a major role in the transport of small solutes,
whereas cilia are paramount for the movement of larger particles.
We also found that the location of cilia, as well as changes to the ventricular geometry,
largely impact solute distribution.
In summary, our work presents a computational
framework that can be applied to other
ventricular systems in the future, together with new concepts
of how cilia distribute molecules and
particles within the brain ventricles.

>>>>

% I2: What do we know about the problem and the topic?
The flow of CSF in the ventricular system is driven by multiple
physiological processes, including via the beating motion of motile
cilia, CSF secretion by the choroid plexus, and pulsatile pressure gradients generated by the
cardiac cycle, respiration, neural activity, and \mer{bodily}
movements~\cite{Brodal2016, del2010ependymal, Ringers2020Role, Vinje2019RespiratoryMeasurements,
Kurtcuoglu2005ComputationalSystem, Olstad2019CiliaryDevelopment,
macaulay2022cerebrospinal, mestre2018flow}.
Cilia are hair-like structures found in a vast diversity of
biological organisms and in several human organs,
including the brain~\cite{DGama2025MotileBrain,
Fulford1986Muco-ciliaryLung,
Pacherres2022CiliaryProduction, Jahn1972LocomotionProtozoa,
Brennen1977FluidFlagella, Ringers2020Role, spassky2013motile,
Pellicciotta2020Cilia, mitchell2007evolution,
Tsukita2012CoordinatedFeet, BLAKE1974MechanicsMotion,
Reiten2017Motile-Cilia-MediatedComputations, Thouvenin2020OriginCanal}. 
The ventricular surface of the brain, known as the ependyma, is lined by
motile cilia that beat in a coordinated pattern at frequencies of
10--40 Hz to mediate fluid flow~\cite{mitchell2007evolution,
Ringers2020Role, spassky2013motile, Ringers2023NovelEpithelia, roth2025structure}. 
Given their microscopic size, ranging from
5 to 15 microns, cilia contribute primarily to flow near the
ependyma in the human brain~\cite{Siyahhan2014FlowVentricles, Olstad2019CiliaryDevelopment,
Ringers2020Role, Faubel2016Cilia-basedVentricles}. In contrast, cardiac,
respiratory and delta waves at lower frequencies induce pulsatile flow
at the larger scales in the brain ventricular system, subarachnoid space, and
subarachnoid perivascular
spaces~\cite{Vinje2019RespiratoryMeasurements, eide2024functional,
Causemann2025, fultz2019coupled, mestre2018flow, daversin2020mechanisms}.

% MER: Note, we do not write "endeavors" nor "delve". (GenAI does.)

% MER: Note "Failed to identify" indicates that someone tried but
% failed. It is a very effective way of pissing said authors (often
% reviewers or potential future collaborators) off if they just didn't
% try/that was not their target.

% I3: What do we not know about the problem and what do we want to
% know? What are the precise research questions we want to address?
% Refer to specific studies and what specifically we do not know.
To date, it remains poorly understood how motile cilia contribute to 
the overall fluid flow and solute transport within the brain.
Experimental approaches face technical challenges: the microscopic size and relatively 
high beating frequencies of the cilia~\cite{ocallaghan2012analysis, mahuzier2018ependymal,
Jeong2022MeasurementTelencephalon, Olstad2019CiliaryDevelopment, spassky2013motile, DGama2021Diversity},
as compared to other fluid dynamic processes in the brain that are slower and span larger
spatial scales~\cite{Vinje2019RespiratoryMeasurements, eide2024functional, fultz2019coupled, mestre2018flow},
make cilia difficult to monitor in vivo in humans.
Until now, most studies have analyzed one physical mechanism at a time, especially in mammals.
For instance, studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles~\cite{Faubel2016Cilia-basedVentricles, Sawamoto2006NewBrain, ibanez2003beat};
however, they \mer{did not study} the role of other contributions to CSF
flow and the accompanying transport, such as pressure
gradients associated with respiration and cardiac pulsations. Moreover, while MRI technology
can measure CSF motion in humans~\cite{feinberg1987human, battal2011cerebrospinal,
yamada2015current}, it cannot detect the microscopic
movements of cilia to understand their impact in the large human
brain~\cite{Siyahhan2014FlowVentricles}. 
These technical limitations of studying cilia-mediated flow and transport make 
mathematical modeling and computational investigations an attractive complement to experimental
studies. Previous simulation studies have focused on modeling the interaction
between single or several cilia and surrounding
fluids~\cite{Guo2020SimulatingGeometries, Ruvalcaba2021NumericalTree, Smith2009MathematicalFluids,
Cui2019NumericalMethod, Cui2022AFlow}.
Others considered cilia-mediated flow in the brain~\cite{Siyahhan2014FlowVentricles,
Yoshida2022EffectVentricles, Salman2022ComputationalEmbryo, Thouvenin2020OriginCanal}, but also 
other applications such as ciliary transport of mucous and particles in
airways~\cite{Fulford1986Muco-ciliaryLung, Ramirez-SanJuan2020Multi-scaleArrays, roth2025structure},
propulsion and locomotion of ciliated organisms~\cite{BLAKE1974MechanicsMotion,
Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella}, and
oxygen regulation in coral reefs~\cite{Pacherres2022CiliaryProduction}.
These simulation studies have provided valuable insights on how cilia mediate complex
biological flows, and how cilia may contribute to tissue physiology.
However, we still lack a comprehensive understanding of how
cilia contribute to CSF motion and solute transport within the brain.
This is of particular importance since disorders such as hydrocephalus are observed
in animal models and humans with deficient motile cilia~\cite{Afzelius2004CiliaRelatedDiseases,
Olstad2019CiliaryDevelopment, Mitchison2017Motile, Ibanez-Tallon2002LossHydrocephalus},
indicating that the cilia play a crucial role for healthy brain function.

% I4: How do we address the problem and what do we find out?
In this study, we combined computational and experimental approaches to investigate
the role of motile cilia in intraventricular CSF flow and transport.
We model the flow and transport by partial differential equations and solve
them numerically with finite element methods to compute flow patterns and
solute concentration dynamics. A geometry representing the zebrafish larval brain ventricles
was used, for which we have detailed knowledge of cilia properties and CSF motion velocities
and frequencies~\cite{Olstad2019CiliaryDevelopment}. 
Notably, this early developmental stage allows us to simplify our model
by omitting CSF secretion, which starts at later stages~\cite{Jeong2024TheZebrafish},
while still being well-conserved with mammals~\cite{DGama2025MotileBrain,
Olstad2019CiliaryDevelopment, Ringers2020Role, DGama2021Diversity, jurisch2020radial}. 
The computational model is validated against experimental data by monitoring in vivo
the dynamics of a photoconvertible fluorescent protein Dendra2
expressed within the brain ventricles.
Our results show that diffusion plays a major role in the transport of small solutes,
whereas cilia are paramount for the movement of larger particles.
We also found that the location of cilia, as well as changes to the ventricular geometry,
largely impact solute distribution.
In summary, our work presents a computational
framework that can be applied to other
ventricular systems in the future, together with new concepts
of how cilia distribute molecules and
particles within the brain ventricles.
